ANALYSIS OF PROSPER LOAN DATA by CHUNDI HIMAKIRAN KUMAR
========================================================

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

```{r packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

install.packages('ggplot2',contriburl = "")

library(ggplot2)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
ld_original <- read.csv("prosperLoanData.csv")
str(ld_original)
# The original prosper loan data has 81 variables.
# we shall use data only for selected 16 variables
ld <- ld_original[,c('CreditGrade','Term','LoanStatus','BorrowerAPR',
'ProsperScore','ListingCategory..numeric.','BorrowerState','Occupation','EmploymentStatus'
,'EmploymentStatusDuration','IsBorrowerHomeowner','IncomeRange','LoanNumber',
'LoanOriginalAmount','LoanOriginationDate','LoanOriginationQuarter')]

```

> **Prosper Loan Dataset**:  The original Prosper Loan Dataset contains 
information of loans cetgorized across 81 variables. For the purpose of this 
analysis we are subsetting this dataset to analyse the data pertaining to 16
variables namely 'CreditGrade','Term','LoanStatus','BorrowerAPR',
'BorrowerRate','ListingCategory..numeric.','BorrowerState','Occupation',
'EmploymentStatus','EmploymentStatusDuration','IsBorrowerHomeowner',
'IncomeRange','LoanNumber','LoanOriginalAmount','LoanOriginationDate' and
'LoanOriginationQuarter'.



# Understanding the structure of the dataset

```{r echo=FALSE, Data_Structure }
str(ld)
```


> **Understanding the range of values for each column**: 

> ***CreditGrade***:

```{r  CreditGrade}
levels(ld$CreditGrade)

```
> ***Term in Months***:

```{r  Term}
range(ld$Term)

```

> ***Term in Months Distribution***:

```{r  Term Distribution}
table(ld$Term)

```

*** Converting Term to factors ***
```{r Term to factors}
ld$Term <- factor(ld$Term, levels=c(12,36,60), ordered=TRUE)

```


> ***Loan Status***:

```{r  Loan Status}
levels(ld$LoanStatus)

```


> ***Loan Status Distribution***:

```{r  Loan Status Distribution}
table(ld$LoanStatus)

```

> ***Borrower APR***:

```{r  Borrower APR}
ld_without_NA <- subset(ld, !is.na(ld$BorrowerAPR))
range(ld_without_NA$BorrowerAPR)

```


> ***Loan Original Amount***:

```{r  Loan Original Amount}
range(ld$LoanOriginalAmount)

```

> ***Employment status duration in months***:

```{r  Employment status duration}
ld_without_NA <- subset(ld, !is.na(ld$EmploymentStatusDuration))
range(ld_without_NA$EmploymentStatusDuration)

```

> ***Prosper score for Loans***:

```{r  Prosper Score}
ld_without_NA <- subset(ld, !is.na(ld$ProsperScore))
range(ld_without_NA$ProsperScore)

```

```{r  Prosper Score distribution}
table(ld$ProsperScore)

```



# Univariate Plots Section




```{r Term Histogram}
ggplot(aes(x = Term), data = ld) +
  geom_bar() +
  ggtitle("Term Histogram")

```

** Term distribution of loans **

```{r summary term}
summary(ld$Term)
```


It is observed that all the loans are either of 12,36 or 60 months duration. The maximum number of loans have a term of 36 
months followed by 60 month term loans.

```{r CreditGrade Histogram}
ggplot(aes(x = CreditGrade), data = ld) +
  geom_histogram(stat = "count") +
  ggtitle("CreditGrade Histogram")

```

```{r summary Credit Grade}
summary(ld$CreditGrade)
```

It is observed that max loans have no credit grade assigned. Among the rest 'C' Credit Grade loans are marginally 
higher than the rest grades.

```{r LoanStatus Histogram}
ggplot(aes(x = LoanStatus), data = ld) +
  geom_histogram(stat = "count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("LoanStatus Histogram")

```

```{r summary Loan Status}
summary(ld$LoanStatus)
```

It is observed that approximately 5000 loans have defaulted status and maximum loans in the dataset are current. The
pst due loans are comparitveley very less.

```{r Borrower APR Histogram}
ggplot(aes(x = BorrowerAPR), data = ld) +
  geom_histogram(binwidth = 0.005) +
  scale_x_continuous(breaks = seq(0,0.5,0.05)) +
  ggtitle("BorrowerAPR Histogram")

```

```{r summary Borrower APR}
summary(ld$BorrowerAPR)
```

It is observed that maximum loans have been issued at an APR of 36%.

```{r EmploymentStatus Histogram}
ggplot(aes(x = EmploymentStatus), data = ld) +
  geom_histogram(stat = "count") +
  ggtitle("EmploymentStatus Histogram")

```

```{r summary Employment Status}
summary(ld$EmploymentStatus)
```

It is observed that maximum loans have been issued to Employed people and very few loans have been issued to 
not employed and part time people.

```{r IncomeRange Histogram}
ggplot(aes(x = IncomeRange), data = ld) +
  geom_histogram(stat = "count") +
  ggtitle("IncomeRange Histogram") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r summary Income Range}
summary(ld$IncomeRange)
```

The histogram is a perfect bell curve and shows that max loans have been given to the middle income groups with 
$25K-49K and $50K-74.99K and the loans fall off on either side as the income rises and falls.

```{r Loan Original Amount Histogram}
ggplot(aes(x = LoanOriginalAmount), data = ld) +
  geom_histogram(binwidth = 1000) +
  scale_x_continuous(breaks = seq(0,35000,2500)) +
  xlab("LoanOriginalAmount(USD)")
  ggtitle("LoanOriginalAmount Histogram")

```

```{r summary Loan Original Amount}
summary(ld$LoanOriginalAmount)
```

The most frequently sanctioned Loan amounts seems to be 4K$, 10K and 15K dollars. There are also loans 
sanctioned for 35K dollars.

```{r Listing Category Histogram}
ggplot(aes(x = ListingCategory..numeric.), data = ld) +
  geom_histogram(stat = "count") +
  scale_x_continuous(breaks = seq(0,20,1)) +
  ggtitle("ListingCategory Histogram")

```

```{r summary Listing Category}
summary(ld$ListingCategory..numeric.)
```

Maximum loans have been sanctioned under listing catgeory 1(Debt Consolidation) followd by 0(Not Available) and 7(Other).
Very few loans have been sanctioned to 8 - Baby&Adoption, 9 - Boat, 10 - Cosmetic Procedure, 11 - Engagement Ring, 
12 - Green Loans followed by 18 - Taxes, 19 - Vacation, 20 - Wedding Loans.

```{r Borrower State Histogram}
ggplot(aes(x = BorrowerState), data = ld) +
  geom_histogram(stat = "count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("BorrowerState Histogram")

```

```{r summary BorrowerState}
summary(ld$BorrowerState)
```

Max lons have been sanctioned to people from state of California followed by approximatley equal number of loans 
to people from Florida, Illinois, New York and Texas.

```{r Employment Status Duration Histogram}
ggplot(aes(x = EmploymentStatusDuration), data = ld) +
  geom_histogram(binwidth=30) +
  scale_x_continuous(breaks = seq(0,600,30)) +
  ggtitle("EmploymentStatusDuration Histogram")

```

```{r summary Employment Status Duration}
summary(ld$EmploymentStatusDuration)
```

It is observed that max loans have been given to people who have only spent a month or two in their employment status and there is a clear downward march of loans given out decreasing as the employment status duration increases.



```{r Prosper Score Histogram}
ld$ProsperScore <- factor(ld$ProsperScore)
ggplot(aes(x = ProsperScore), data = ld_without_NA) +
  geom_bar() +
  scale_x_continuous(breaks = seq(1,11,1)) +
  ggtitle("ProsperScore Histogram")

```

```{r summary Propser Score}
summary(ld$ProsperScore)
```

It is observed that max loans have a prosper score of 4,6 and 8 while there are comparitevely very few loans having the worst score of 1 and best score of 11. 


# Univariate Analysis

### Main features of interest in your dataset?

Based on the above analysis the PropserScore, BorrowedAPR and the Loan Amount
seem to be features of interest


### Other features in the dataset which will help support investigation into 
### the above feature(s) of interest?

IsBorrowerHomeowner, Term and EmploymentStatusDuration are the features which
could support the investigation.

### Unusual Distributions..
The employment status duration histogram was unexpected as it showed that the
loans granted to people decreased as their duration in the respective employment
increased. It perhaps could be attributed to the fact that max loans were given
out to new employees upon getting a job within the first two months.


# Bivariate Plots Section

```{r Prosper Score Histogram colored}
ggplot(aes(x = ProsperScore), data = ld_without_NA) +
  geom_histogram(aes(color = IsBorrowerHomeowner)) +
  ggtitle("PropserScore Histogram Colored by Home Owners")

```

We see that maximum loans have been given to non home owners as compared to home owners across all the prosper score ratings.

```{r Term values distribution}
table(ld$Term)
```


```{r echo=FALSE, BorrowerAPR_Term}
ggplot(aes(x = Term, y = BorrowerAPR), data = ld)+
  geom_boxplot() +
  ggtitle("BorrowerAPR vs Term")
```

There are only three values of term ie 12,36 and 60 months.We can see that the median BorrowerAPR is approximately the same for all three terms.
But the range of BorrowerAPR values is much larger for 36 Months Term Loans than others.


```{r echo=FALSE, BorrowerAPR_LoanOriginalAmount}
ggplot(aes(x = LoanOriginalAmount, y = BorrowerAPR), data = ld)+
  geom_jitter(alpha=1/20) +
  xlab("LoanOriginalAmount(USD)") +
  ggtitle("BorrowerAPR vs LoanOriginalAmount")
```

As the Loan amount increases the Borrower APR moves to a range between 0.1 to 0.2 from 0.05 to 0.4.
There seems to be a negative corelation between Borrower APR and Loan Amount.

```{r echo=FALSE, BorrowerAPR_EmploymentStatusDuration }
ggplot(aes(x = EmploymentStatusDuration, y = BorrowerAPR), data = ld) +
  geom_jitter(alpha = 1/40) +
  ggtitle("BorrowerAPR vs EmploymentStatusDuration")

```

We want to improve the above plot by coverting the employment status duration from days to months.

```{r  BorrowerAPR_EmploymentStatusDuration_Inyears }
ggplot(aes(x = (EmploymentStatusDuration%%12), y = BorrowerAPR), data = ld) +
  geom_jitter(alpha=1/40) +
  ggtitle("BorrowerAPR vs EmploymentStatusDuration in Months")

```

We now see that Employment status duration month wise has a similar range of APR for all months. However we also observe
that many loans with varying values of employment status duration have an APR whci is 0.35-0.36 percentage points.

```{r echo=FALSE, LoanAmount_EmploymentStatusDuration }
ggplot(aes(x = EmploymentStatusDuration, y = LoanOriginalAmount), data = ld) +
  geom_jitter(alpha = 1/40) +
  ylab("LoanOriginalAmount(USD)") +
  ggtitle("LoanOriginalAmount vs EmploymentStatusDuration")

```

Here also we shall try to change the value of EmploymentStatus duration from days to months to revisit this graph

```{r echo=FALSE, LoanAmount_EmploymentStatusDuration2 }
ggplot(aes(x = (EmploymentStatusDuration%%12), y = LoanOriginalAmount), data = ld) +
  geom_jitter(alpha=1/40) +
  ylab("LoanOriginalAmount(USD)") +
  ggtitle("LoanOriginalAmount vs EmploymentStatusDuration in months")
  
```

We now see that Employment status duration for varying montha has similar Loan Amount distribution. But we also notice that maximum
loans in each value of Employment status duration have a Loan Amounts of 10K$,15K$, 20K$ and 25K$.

```{r echo=FALSE, Term_LoanAmount }
ggplot(aes(x = Term, y = LoanOriginalAmount), data = ld) +
  geom_boxplot() +
  ylab("LoanOriginalAmount(USD)") +
  ggtitle("LoanOriginalAmount vs Term")

```

We see that median Loan Amount increases with increase in the term distribution.

```{r echo=FALSE, ProsperScore_LoanAmount }
ggplot(aes(x = ProsperScore, y = LoanOriginalAmount), data = ld) +
  geom_boxplot() +
  ylab("LoanOriginalAmount(USD)") +
  ggtitle("LoanOriginalAmount vs ProsperScore")

```

We see that as the PropserScore increases the Median LoanOriginalAmount also tends to increase.

```{r echo=FALSE, ProsperScore_EmploymentStatusDuration}
ggplot(aes(x = ProsperScore, y = EmploymentStatusDuration), data = ld)+
  geom_boxplot() +
  ggtitle("EmploymentStatusDuration vs ProsperScore")
```

The median Employment status duration does not seem to change much with the propser score.

```{r echo=FALSE, ProsperScore_BorrowerAPR}
ggplot(aes(x = ProsperScore, y = BorrowerAPR), data = ld)+
  geom_boxplot() +
  ggtitle("BorrowerAPR vs ProsperScore")
```

We can see that the median Borrower APR decreases with increasing Prosper Score..Hence we would want to further analyze this relationship
as under. We shall group the data by prosper score and try to find the Borrower APR mean for each group and then plot the mean vs the 
Prosper Score. To do this we shall use 'dplyr' package and derive a dataframe by grouping the existing dataframe by ProsperScore.

```{r  BorrowerAPR_Mean_groupby_prosperscore}
library(dplyr)
ld_by_p_score <- group_by(ld, ProsperScore)
ld.BorrowerAPR_by_P_Score_mean <- summarise(ld_by_p_score,
                          BorrowerAPR_Mean = mean(BorrowerAPR),
                          BorrowerAPR_Median = median(BorrowerAPR),
                          LoanAmount_Mean = mean(LoanOriginalAmount),
                          n = n())
ld.BorrowerAPR_by_P_Score_mean <- arrange(ld.BorrowerAPR_by_P_Score_mean, ProsperScore)

```

```{r  BorrowerAPR_Mean_ProsperScore}
ggplot(aes(x = ProsperScore, y = BorrowerAPR_Mean), data = ld.BorrowerAPR_by_P_Score_mean)+
  geom_boxplot() +
  ggtitle("BorrowerAPR_Mean vs ProsperScore")

```

As expected we find a very close inverse linear relationship between ProsperScore and the mean Borrower APR.
We shall also try and find a relationship between the ProsperScore and the Loan Amount sanctioned.

```{r  MeanLoanAmount_ProsperScore_Groupby}
ggplot(aes(x = ProsperScore, y = LoanAmount_Mean), data = ld.BorrowerAPR_by_P_Score_mean)+
  geom_boxplot() +
  ylab("LoanOriginalAmount_Mean(USD)") +
  ggtitle("LoanAmount_Mea vs ProsperScore")
```

 We see that there is an increasing linear relationship between the mean Loan Amount and the Prosper Score.



 

# Bivariate Analysis


###  Relationships of Interest

By grouping the loans using the ProsperScore and then finding the BorrowerAPR mean of each group we were able to plot and see the
relationship between the ProsperSCore and BorrowerAPR mean as well as ProsperScore and LoanOriginalAmount. We saw that highly rated loans
had more Loan Amount Sanctioned and lesser Borrower APR.




# Multivariate Plots Section

```{r  BorrowerAPR_PropserScore}
ggplot(aes(x = ProsperScore, y = BorrowerAPR), data = ld)+
  geom_jitter(alpha=1/40,aes(color = IsBorrowerHomeowner))+
  geom_smooth() +
  ggtitle("BorrowerAPR vs ProsperScore Colored by HomeOwner")

```

We see that across the prosper score ratings People who are not Home owners(red color) tend to have higher Borrower APR. We also
see that as the propser score increases (gets better) the Borrower APR goes down.

```{r echo=FALSE, BorrowerAPR_LoanOriginalAmount_Colored_By_Homeowner}
ggplot(aes(x = BorrowerAPR, y = LoanOriginalAmount), data = ld)+
  geom_line(aes(color = IsBorrowerHomeowner)) +
  ylab("LoanOriginalAmount(USD)") +
  ggtitle("LoanOriginalAmount vs BorrowerAPR Colored by HomeOwner")
```

We wanted to further understand the relationship between Loan Original Amount and Borrower APR. By colouring the data using
the HomeOwner variable we see that maximum loans having APR from 0.05 to 0.4 have been given to the HomeOwners.
Loans with Loan Amounts more than 0.4 are mostly given to People without home ownership.

```{r echo=FALSE, ProsperScore_LoanAmount_ColoredByHomeowner }
ggplot(aes(x = ProsperScore, y = LoanOriginalAmount), data = ld) +
  geom_jitter(alpha=1/40,aes(color = IsBorrowerHomeowner)) +
  geom_smooth() +
  ylab("LoanOriginalAmount(USD)") +
  ggtitle("LoanOriginalAmount vs ProsperScore Colored by HomeOwner")

```

We wanted to extend the analysis of ProsperScore and the LoanOriginalAmount using the home owner variable.
From the graph above we see that People who are not home owners form the majority of loans that have lesser loan original amounts
as compared to loans of home owners.Loans upward of 10K$ are mostly concentrated with Home owners. It is also seen that Loans
with prosper score of 9 and above are predominantly of Home owners.

# Multivariate Analysis

We have used the IsBorrowerHomeowner variable to extend our analysis between the ProsperScore , BorrowerAPR against the
LoanOriginalAmount. The aim was to see if the variable IsBorrowerHomeOwner would give us any indication about the interplay 
of these three variables.From the above two graphs we can conclude that Loans given to Homeowners have better APR, Larger LoanAmounts and better prosper scores as compared to non home owners.



------

# Final Plots and Summary


### Plot One
```{r  MeanLoanAmount_ProsperScore}
ggplot(aes(x = ProsperScore, y = LoanAmount_Mean), data = ld.BorrowerAPR_by_P_Score_mean)+
  geom_boxplot() +
  ylab("LoanOriginalAmount_Mean(USD)") +
  ggtitle("LoanOriginalAmount_Mean vs ProsperScore")
```


### Description One

The mean Loan Amount of loans grouped by ProsperScore has a clear approximate
increasing linear relationship with the ProsperScore. It can be understood 
logically that highly rated loans have a larger Loan Amount than loans with poor
rating.


### Plot Two

```{r  BorrowerAPR_Mean_PropserScore}
ggplot(aes(x = ProsperScore, y = BorrowerAPR_Mean), data = ld.BorrowerAPR_by_P_Score_mean)+
  geom_boxplot() +
  ggtitle("BorrowerAPR_Mean vs ProsperScore")
```


### Description Two

As an inverse of the previous graph we can see here that the APR mean falls
with increasing Prosper Score. Again this can be understood in a way that the
loans with higher Prosper SCore have a mean APR lower than the loans with poorer
Prosper Scores.



### Plot Three

```{r  BorrowerAPR_PropserScore_ColoredByHomeOwner}
ggplot(aes(x = ProsperScore, y = BorrowerAPR), data = ld)+
  geom_jitter(alpha=1/40,aes(color = IsBorrowerHomeowner))+
  geom_smooth() +
  ggtitle("BorrowerAPR vs ProsperScore Colored by HomeOwner")

```

### Description Three
We see that across the prosper score ratings People who are not Home owners(red color) tend to have higher Borrower APR. We also
see that as the propser score increases (gets better) the Borrower APR goes down.

------

# Reflection

It was a difficult process to analyze this dataset. The intial difficulty lay
in trying to guess the interesting variables for analysis from a total of 81
variables. The distributions of the variables helped understand the nature 
of the data and the distribution which in turn help guess the pair of variables
that could seem to have some relationship between them. After narrowing down to 
the interesting relationship between Borrowed APR and Loan Amount it was decided
to explore this relationship further using the ProsperScore variable.
Finally we succeeded in establishing that both the Loan Amount and Borrowed APR
varied in a linear fashion compared to the ProsperScore variable.

Further work can include fitting a model to Final Plots One and Two..